---
title: "array example"
author: "Benjamin Galuardi"
date: "Monday, December 14, 2015"
output: html_document
---

Populates top to bottom, left to right
```{r}
a = array(1:12, dim = c(3,4))
a 
```

use Matrix. Populates by column unless specified
```{r}
# By Column
b = matrix(1:12, nrow = 3, ncol = 4)
b 
```
Populate by row: Populates left to right, top to bottom
```{r}
# By Row
b = matrix(1:12, nrow = 3, ncol = 4, byrow=T)
b 
```

Now flip across the y axis
```{r}
s = b[nrow(b):1,]
s
```
Now, populates left to right, bottom to top. This is typically how satellite data is presented. Bottom left as (0,0)

What if you start with satellite data and make an array?
---
```{r}
as.array(s)
```

If you transpose an array with different dimensions, things can get messed up
```{r}
# Transpose
t(s)
```


more tricks
---
designate the total size of the matrix/array/list before populating. This uses memory more efficiently and is faster.

```{r}
z = array(0, dim=c(3,4,2))
z
```

now populate the array

```{r}
z[,,1] = a
z[,,2] = b
z
```


